---
title: "Thesis Data Analysis"
author: "Charles Beckler"
date: "February 5, 2019"
output:
  pdf_document:
    toc: yes
  html_notebook:
    toc: yes
---

The purpose of this analysis is to determine whether transgender men (FTM) are experiencing the phenomena of physical gender dysphoria different from transgender women (MTF). Exploratory factor analyses are performed to find if the genders present different factors, indicating differences in subjective experience. Confirmatory factor analyses test if my exploratory model fits better than the model in the established literature. Finally, two group ANOVAs test whether medical hormone treatment affects mean values within genders.

#Loading required packages

```{r message=FALSE}
library(psych)
library(GPArotation) 
library(foreign)
library(lavaan)
library(sjstats)
library(plyr)
library(dplyr)
```

#Importing data

```{r message=FALSE}
setwd("C:/Users/charl/OneDrive/Documents/School/Fall Thesis/final data")

ftm <- read.spss("FTM hormone.sav", header = TRUE, use.value.labels = FALSE)

mtf  <- read.spss("MTF hormone.sav", header = TRUE, use.value.labels = FALSE)
```

#Creating usable data frames

```{r}
ftm2 <- data.frame(ftm$TransitionStatus,ftm$p1,ftm$p2,ftm$p3,ftm$p4,ftm$p5,ftm$p6,ftm$p7,ftm$s1,ftm$s2,ftm$s3,ftm$s4,
                   ftm$s5,ftm$s6,ftm$s7,ftm$s8,ftm$s9,ftm$s10,ftm$s11,ftm$s12,ftm$s13,ftm$n1,ftm$n2,ftm$n3,ftm$n4,
                   ftm$n5,ftm$n6,ftm$n7,ftm$n8,ftm$n9,ftm$n10)

mtf2 <- data.frame(mtf$TransitionStatus,mtf$p1,mtf$p2,mtf$p3,mtf$p4,mtf$p5,mtf$p6,mtf$s1,mtf$s2,mtf$s3,mtf$s4,mtf$s5,
                   mtf$s6,mtf$s7,mtf$s8,mtf$s9,mtf$s10,mtf$s11,mtf$s12,mtf$s13,mtf$s14,mtf$n1,mtf$n2,mtf$n3,mtf$n4,
                   mtf$n5,mtf$n6,mtf$n7,mtf$n8,mtf$n9,mtf$n10)
```

#Validity Checks

##Creating data frames for each scale of measurement

```{r}
ftmPrimary <- data.frame(ftm2$ftm.p1, ftm2$ftm.p2, ftm2$ftm.p3, ftm2$ftm.p4, ftm2$ftm.p5, ftm2$ftm.p6, ftm2$ftm.p7)

ftmSecondary <- data.frame(ftm2$ftm.s1, ftm2$ftm.s2, ftm2$ftm.s3, ftm2$ftm.s4, ftm2$ftm.s5, ftm2$ftm.s6, ftm2$ftm.s7,
                           ftm2$ftm.s8, ftm2$ftm.s9, ftm2$ftm.s10, ftm2$ftm.s11, ftm2$ftm.s12, ftm2$ftm.s13)

ftmNeutral <- data.frame(ftm2$ftm.n1, ftm2$ftm.n2, ftm2$ftm.n3, ftm2$ftm.n4, ftm2$ftm.n5, ftm2$ftm.n6, ftm2$ftm.n7,
                         ftm2$ftm.n8, ftm2$ftm.n9, ftm2$ftm.n10)

mtfPrimary <- data.frame(mtf2$mtf.p1, mtf2$mtf.p2, mtf2$mtf.p3, mtf2$mtf.p4, mtf2$mtf.p5, mtf2$mtf.p6)

mtfSecondary <- data.frame(mtf2$mtf.s1, mtf2$mtf.s2, mtf2$mtf.s3, mtf2$mtf.s4, mtf2$mtf.s5, mtf2$mtf.s6, mtf2$mtf.s7,
                           mtf2$mtf.s8, mtf2$mtf.s9, mtf2$mtf.s10, mtf2$mtf.s11, mtf2$mtf.s12, mtf2$mtf.s13,
                           mtf2$mtf.s14)

mtfNeutral <- data.frame(mtf2$mtf.n1, mtf2$mtf.n2, mtf2$mtf.n3, mtf2$mtf.n4, mtf2$mtf.n5, mtf2$mtf.n6, mtf2$mtf.n7,
                         mtf2$mtf.n8, mtf2$mtf.n9, mtf2$mtf.n10)
```

##Cronbach's Alpha

###FTM

```{r}
alpha(ftmPrimary)
```

Cronbach's alpha for the FTM version of the Primary Sexual Characteristics scale is only 0.67, which is a weak alpha.

```{r}
alpha(ftmSecondary)
```

Cronbach's alpha for the FTM Secondary Sexual Characteristics Scale is 0.83, an acceptable alpha.

```{r}
alpha(ftmNeutral)
```

Cronbach's alpha for the FTM Neutral Characteristics scale is 0.65, a weak alpha.


Already we can see that the FTM version of the measurement instrument has significant issues.

###MTF

```{r}
alpha(mtfPrimary)
```

Cronbach's alpha for the MTF Primary Sexual Characteristics scale is 0.62, a weak alpha.

```{r}
alpha(mtfSecondary)
```

Cronbach's alpha for the MTF Secondary Sexual Characteristics scale is 0.84, an acceptable alpha.

```{r}
alpha(mtfNeutral)
```

Cronbach's alpha for the MTF Neutral Characteristics scale is 0.73, a marginally acceptable alpha.


We can see there are significant problems with the MTF version of the measurement instrument as well.

#Exploratory Factor Analysis

In this analysis I will use the observed data to determine which items are clustering together in factors.

##FTM

```{r}
ftm3 <- subset(ftm2, select = -1)

fa.parallel(ftm3, fm = 'minres', fa = 'fa')
```

I first use the subset function to create a new data frame without the hormone status variable, which is not needed in this analysis but will be used later.

Horn's parallel analysis tells me that the number of factors present in the observed data set is 8.

```{r}
ftmEFA <- fa(ftm3, nfactors = 8, rotate = "varimax", fm="minres")

print(ftmEFA)
```

The RMSEA for this model is 0.035, CI (0.024, 0.041). This demonstrates the model is statistically sound.

```{r}
print(ftmEFA$loadings,cutoff = 0.3)
```

This read out shows me which items are clustering together to make a factor within the data. I have excluded any correlations below 0.3, so this is only items that are correlated moderately or stronger. Later, I look at the individual items to determine the qualitative meaning of the factor: interpreting it cohesively.

##MTF

```{r}
mtf3 <- subset(mtf2, select = -1)

fa.parallel(mtf3, fm = 'minres', fa = 'fa')
```

Again, the hormone condition variable is dropped. Horn's parallel analysis tells us that the number of factors present in the data is 5.

```{r}
mtfEFA <- fa(mtf3, nfactors = 5, rotate = "varimax", fm="minres")

print(mtfEFA)
```

The RMSEA of this model is 0.077, CI (0.046, NA), indicating this model is an acceptable fit, but the upper bound confidence interval is having issues due to the small sample size and results must be interpreted with caution.

```{r}
print(mtfEFA$loadings,cutoff = 0.3)
```

This read out shows me which items are clustering together to make a factor within the data. I have excluded any correlations below 0.3, so this is only items that are correlated moderately or stronger. Later, I look at the individual items to determine the qualitative meaning of the factor: interpreting it cohesively.

#Confirmatory Factor Analysis

In this analysis, I compute the RMSEA for the established model to see how it compares to my exploratory model.

##FTM

```{r}
modelftm = '
  Prime =~ ftm.p1 + ftm.p2 + ftm.p3 + ftm.p4 + ftm.p5 + ftm.p6 + ftm.p7
  Second =~ ftm.s1 + ftm.s2 + ftm.s3 + ftm.s4 + ftm.s5 + ftm.s6 + ftm.s7 + ftm.s8 + ftm.s9 + ftm.s10 + ftm.s11 + ftm.s12 + ftm.s13 
  Neutral =~ ftm.n1 + ftm.n2 + ftm.n3 + ftm.n4 + ftm.n5 + ftm.n6 + ftm.n7 + ftm.n8 + ftm.n9 + ftm.n10
'

fitftm = sem(modelftm, data = ftm3)
summary(fitftm, fit.measures = TRUE, standardized = TRUE)
```

I have used structural equation modeling to perform a confirmatory factor analysis on the observed data. The RMSEA is 0.082, CI (0.078, 0.087). This is a considerably worse fit than my exploratory model.

##MTF

```{r}
modelmtf = '
  Prime =~ mtf.p1 + mtf.p2 + mtf.p3 + mtf.p4 + mtf.p5 + mtf.p6
  Second =~ mtf.s1 + mtf.s2 + mtf.s3 + mtf.s4 + mtf.s5 + mtf.s6 + mtf.s7 + mtf.s8 + mtf.s9 + mtf.s10 + mtf.s11 + mtf.s12 + mtf.s13 + mtf.s14 
  Neutral =~ mtf.n1 + mtf.n2 + mtf.n3 + mtf.n4 + mtf.n5 + mtf.n6 + mtf.n7 + mtf.n8 + mtf.n9 + mtf.n10
'

fitmtf = sem(modelmtf, data = mtf3)
summary(fitmtf, fit.measures = TRUE, standardized = TRUE)
```

Again I have used structural equation modeling to perform a confirmatory factor analysis. The RMSEA is 0.109, CI (0.099, 0.118), which is unacceptably large. This is a considerably worse fit than my exploratory model.

#Model Comparison

##FTM

```{r}
modelftmEFA = '
  one =~ ftm.s1 + ftm.s2 + ftm.s3 + ftm.s5 + ftm.s7 + ftm.s10 + ftm.s11 + ftm.n4
  two =~ ftm.s4 + ftm.s6 + ftm.s9
  three =~ ftm.s2 + ftm.s3 + ftm.s7 + ftm.n1 + ftm.n3 + ftm.n8 + ftm.n9
  four =~ ftm.p1 + ftm.p2 + ftm.p3
  five =~ ftm.p6 + ftm.p7 + ftm.s13
  six =~ ftm.p4 + ftm.p5
  seven =~ ftm.s8 + ftm.n7
  eight =~ ftm.n5 + ftm.n10
'
ftmMine = sem(modelftmEFA, data = ftm3)

anova(ftmMine, fitftm)
```

My model fits significantly better.

##MTF

```{r}
modelmtfEFA = '
  one =~ mtf.p4 + mtf.p5 + mtf.s4 + mtf.s7 + mtf.s8 + mtf.s13 + mtf.s14 + mtf.n1 + mtf.n2 + mtf.n3 + mtf.n5 + mtf.n6 + mtf.n7 + mtf.n8 + mtf.n9 + mtf.n10
  two =~ mtf.p6 + mtf.s1 + mtf.s2 + mtf.s3 + mtf.s5 + mtf.s7 + mtf.s8 + mtf.s10 + mtf.s11 + mtf.s12 + mtf.s14 + mtf.n4
  three =~ mtf.p1 + mtf.p2 + mtf.p3
  four =~ mtf.s4 + mtf.s6 + mtf.s9 + mtf.s10 + mtf.n2
  five =~ mtf.p5 + mtf.p6 + mtf.s14 + mtf.n6
'

mtfMine = sem(modelmtfEFA, data = mtf3)

anova(mtfMine, fitmtf)
```

My model fits significantly better.

#Multi-group CFA by transition status

no hormones = 1, hormones = 2

##FTM

```{r}
ftmhormone <- sem(modelftmEFA, data = ftm2, group = "ftm.TransitionStatus", orthogonal = FALSE)

summary(ftmhormone, fit.measures = TRUE)
```

The model still fits well, RMSEA = 0.042, CI (0.034, 0.049), but there does not appear to be a large difference between groups (no HRT chi = 396.9, HRT chi = 387.5).

##MTF

```{r}
modelmtfEFA = '
  one =~ mtf.p4 + mtf.p5 + mtf.s4 + mtf.s7 + mtf.s8 + mtf.s13 + mtf.s14 + mtf.n1 + mtf.n2 + mtf.n3 + mtf.n5 + mtf.n6 + mtf.n7 + mtf.n8 + mtf.n9 + mtf.n10
  two =~ mtf.p6 + mtf.s1 + mtf.s2 + mtf.s3 + mtf.s5 + mtf.s7 + mtf.s8 + mtf.s10 + mtf.s11 + mtf.s12 + mtf.s14 + mtf.n4
  three =~ mtf.p1 + mtf.p2 + mtf.p3
  four =~ mtf.s4 + mtf.s6 + mtf.s9 + mtf.s10 + mtf.n2
  five =~ mtf.p5 + mtf.p6 + mtf.s14 + mtf.n6
  cs6 > 0
  mtf.s6~~c(es6,cs6)*mtf.s6
'

mtfhormone <- sem(modelmtfEFA, data = mtf2, group = "mtf.TransitionStatus", orthogonal = FALSE)

summary(mtfhormone, fit.measures = TRUE)
```

Overall this model does not fit, so there is no difference between groups. I'm getting some error messages but with RMSEA = 0.116 I think it is safe to conclude the model is not working.

#Within gender mean comparisons by hormone treatment

I want to do mean comparisons to inspect *why* there is no difference. I suspect what is happening is that several of the factors are not strongly impacted by hormones, but maybe one or two are. It would be interesting to see which ones are and report on that.a

Hormone treatment is coded as 1 = never on hormones, 2 = currently on hormones.

##FTM

###Creating mean variables

I must first add new variables to my data set with the means of the exploratory factors I found. I must also make Transiton Status a factor instead of numerical.

```{r message = FALSE}
ftm2 <- mutate(ftm2, meanF1 = (ftm.s1+ftm.s2+ftm.s3+ftm.s5+ftm.s7+ftm.s10+ftm.s11+ftm.n4)/8)
ftm2 <- mutate(ftm2, meanF2 = (ftm.s4 + ftm.s6 + ftm.s9)/3)
ftm2 <- mutate(ftm2, meanF3 = (ftm.s2 + ftm.s3 + ftm.s7 + ftm.n1 + ftm.n3 + ftm.n9 + ftm.n10)/7)
ftm2 <- mutate(ftm2, meanF4 = (ftm.p1 + ftm.p2 + ftm.p3)/3)
ftm2 <- mutate(ftm2, meanF5 = (ftm.p6 + ftm.p7 + ftm.s13)/3)
ftm2 <- mutate(ftm2, meanF6 = (ftm.p4 + ftm.p5)/2)
ftm2 <- mutate(ftm2, meanF7 = (ftm.s8 + ftm.n8)/2)
ftm2 <- mutate(ftm2, meanF8 = (ftm.n6 + ftm.n10)/2)

ftm2$ftm.TransitionStatus <- factor(ftm2$ftm.TransitionStatus)

```

###Get count per group

```{r}
#count(ftm2$ftm.TransitionStatus)
```

This works in R but not in the knit for some reason so I just commented it out. It's not that important. No hormone = 243, hormone = 148.

###Mean comparisons

I will run basic two group ANOVAs to find if the two groups differ significantly in mean values. Because I expect to get low p-values with such a large sample size, I will calculate the effect size. Tapply gives us the mean score for each within-gender group. 

```{r}
tapply(ftm2$meanF1,ftm2$ftm.TransitionStatus,mean)
f1<-aov(meanF1~ftm.TransitionStatus, data=ftm2)
summary.lm(f1)
eta_sq(f1)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(ftm2$meanF2,ftm2$ftm.TransitionStatus,mean)
f2<-aov(meanF2~ftm.TransitionStatus, data=ftm2)
summary.lm(f2)
eta_sq(f2)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(ftm2$meanF3,ftm2$ftm.TransitionStatus,mean)
f3<-aov(meanF3~ftm.TransitionStatus, data=ftm2)
summary.lm(f3)
eta_sq(f3)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(ftm2$meanF4,ftm2$ftm.TransitionStatus,mean)
f4<-aov(meanF4~ftm.TransitionStatus, data=ftm2)
summary.lm(f4)
eta_sq(f4)
```

Because p > 0.05, there is no significant difference between groups.

```{r}
tapply(ftm2$meanF5,ftm2$ftm.TransitionStatus,mean)
f5<-aov(meanF5~ftm.TransitionStatus, data=ftm2)
summary.lm(f5)
eta_sq(f5)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(ftm2$meanF6,ftm2$ftm.TransitionStatus,mean)
f6<-aov(meanF6~ftm.TransitionStatus, data=ftm2)
summary.lm(f6)
eta_sq(f6)
```

Because p > 0.05, there is not a significant difference between groups.

```{r}
tapply(ftm2$meanF7,ftm2$ftm.TransitionStatus,mean)
f7<-aov(meanF7~ftm.TransitionStatus, data=ftm2)
summary.lm(f7)
eta_sq(f7)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(ftm2$meanF8,ftm2$ftm.TransitionStatus,mean)
f8<-aov(meanF8~ftm.TransitionStatus, data=ftm2)
summary.lm(f8)
eta_sq(f8)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

Most have small effect sizes but the factor with eta square = 0.22 may be worth reporting on.

**Though most of the ANOVAs came out significant, all the effect sizes are small. It is reasonable to conclude there is not truly a meaningful difference between groups.**

##MTF

###Creating mean value variables

```{r}
mtf2 <- mutate(mtf2, meanF1 = (mtf.p4 + mtf.p5 + mtf.s4 + mtf.s7 + mtf.s8 + mtf.s13 + mtf.s14 + mtf.n1 + mtf.n2 + mtf.n3 + mtf.n5 + mtf.n6 + mtf.n7 + mtf.n8 + mtf.n9 + mtf.n10)/16)
mtf2 <- mutate(mtf2, meanF2 = (mtf.p6 + mtf.s1 + mtf.s2 + mtf.s3 + mtf.s5 + mtf.s7 + mtf.s8 + mtf.s10 + mtf.s11 + mtf.s12 + mtf.s14 + mtf.n4)/12)
mtf2 <- mutate(mtf2, meanF3 = (mtf.p1 + mtf.p2 + mtf.p3)/3)
mtf2 <- mutate(mtf2, meanF4 = (mtf.s4 + mtf.s6 + mtf.s9 + mtf.s10 + mtf.n2)/5)
mtf2 <- mutate(mtf2, meanF5 = (mtf.p5 + mtf.p6 + mtf.s14 + mtf.n7)/4)

mtf2$ftm.TransitionStatus <- factor(mtf2$mtf.TransitionStatus)
```

###Getting a count per group

```{r}
#count(mtf2$mtf.TransitionStatus)
```

No hormone = 45, hormone = 59.

###Mean comparisons

```{r}
tapply(mtf2$meanF1,mtf2$mtf.TransitionStatus,mean)
m1<-aov(meanF1~mtf.TransitionStatus, data=mtf2)
summary.lm(m1)
eta_sq(m1)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(mtf2$meanF2,mtf2$mtf.TransitionStatus,mean)
m2<-aov(meanF2~mtf.TransitionStatus, data=mtf2)
summary.lm(m2)
eta_sq(m2)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

```{r}
tapply(mtf2$meanF3,mtf2$mtf.TransitionStatus,mean)
m3<-aov(meanF3~mtf.TransitionStatus, data=mtf2)
summary.lm(m3)
eta_sq(m3)
```

Because the p-value > 0.05, there is no significant difference.

```{r}
tapply(mtf2$meanF4,mtf2$mtf.TransitionStatus,mean)
m4<-aov(meanF4~mtf.TransitionStatus, data=mtf2)
summary.lm(m4)
eta_sq(m4)
```

Because the p-value > 0.05, there is no significant difference.

```{r}
tapply(mtf2$meanF5,mtf2$mtf.TransitionStatus,mean)
m5<-aov(meanF5~mtf.TransitionStatus, data=mtf2)
summary.lm(m5)
eta_sq(m5)
```

Because p < 0.05, the difference is significant for this factor, with the hormone group experiencing lower scores.

This is a small effect size.

Again, the factor with eta square = 0.22 might be worth reporting on.

**This concludes the analyses.**